---
title: "Class 9: Candy Mini-Project"
author: "Mayla (PID: A19136116)"
format: pdf
toc: true
---

## 1. Background

Today we are taking a detor to analyze a fun dataset (that we have more intrinsic insight into) with the most useful analysis method we have learned thus far - Principal Component Analysis (PCA)

## 2.Importing candy data

```{r}
candy_file <- "candy-data.csv"

candy = read.csv("https://raw.githubusercontent.com/fivethirtyeight/data/master/candy-power-ranking/candy-data.csv",
                 row.names = 1)
head(candy)
```

#### **2.1 What is in the dataset?**

> **Q1. How many different candy types are in this dataset?**

```{r}
nrow(candy)
```

> **Q2. How many fruity candy types are in the dataset?**

```{r}
sum(candy$fruity)
```

#### 2.2 What is your favorite candy

```{r}
candy["Twix", ]$winpercent
```

```{r}
library(dplyr)

candy |> 
  filter(row.names(candy)=="Twix") |> 
  select(winpercent)
```

> **Q3. What is your favorite candy (other than Twix) in the dataset and what is it’s winpercent value?**

```{r}
candy["Hershey's Milk Chocolate", ]$winpercent
```

> **Q4. What is the `winpercent` value for “Kit Kat”?**

```{r}
candy["Kit Kat", ]$winpercent
```

> **Q5. What is the `winpercent` value for “Tootsie Roll Snack Bars”?**

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```

\#**Side Note**

```{r}
library("skimr")
skimr::skim(candy)
```

From your use of the skim() function use the output to answer the following:

> **Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?**

```{r}
summary(candy)
```

Yes — winpercent, sugarpercent, and pricepercent are continuous, while most other variables are binary (0/1).

> **Q7. What do you think a zero and one represent for the candy\$chocolate column?**

```{r}
skim("chocolate")
```

1 = contains chocolate, 0 = does not contain chocolate.

## 3.Exploratory Analysis

> **Q8. Plot a histogram of winpercent values using both base R and ggplot2.**

Base R

```{r}
hist(candy$winpercent,
     main="Histogram of Win Percent",
     xlab="Win Percent",
     col="lightblue")
```

ggplot2

```{r}
library(ggplot2)

ggplot(candy, aes(x=winpercent)) +
  geom_histogram(binwidth=5, fill="pink", color="black") +
  labs(title="Histogram of Win Percent", x="Win Percent")
```

> **Q9. Is the distribution of winpercent values symmetrical?**

No, the distribution of winpercent values is not perfectly symmetrical; it appears slightly skewed.

> **Q10. Is the center of the distribution above or below 50%?**

```{r}
mean(candy$winpercent)
```

The center of the distribution is slightly above 50%.

> **Q11. On average is chocolate candy higher or lower ranked than fruit candy?**

```{r}
mean(candy$winpercent[candy$chocolate == 1])
mean(candy$winpercent[candy$fruity == 1])
```

Chocolate is higher ranked than fruit candy.

> **Q12. Is this difference statistically significant?**

```{r}
t.test(winpercent ~ chocolate, data = candy)
```

Yes, the difference is statistically significant, as the p-value from the t-test is less than 0.05.

## 4. Overall Candy Rankings

> **Q13. What are the five least liked candy types in this set?**

```{r}
head(candy[order(candy$sugarpercent),], n=5)
```

The five least liked candies are: Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, and Jawbusters.

> **Q14. What are the top 5 all time favorite candy types out of this set?**

```{r}
head(candy[order(candy$winpercent, decreasing = TRUE), ], 5)
```

The top five most popular candies are: Reese’s Peanut Butter Cup, Reese’s Miniatures, Twix, Kit Kat, and Snickers.

> **Q15. Make a first barplot of candy ranking based on `winpercent` values.**

```{r, fig.height=15, fig.width=10}
library(ggplot2)

ggplot(candy) +
  aes(winpercent, rownames(candy)) +
  geom_col()
```

> **Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?**

```{r, fig.height=15, fig.width=10}
library(ggplot2)

ggplot(candy) +
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col()
```

#### 4.0.1 Time to add some useful color

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[candy$bar == 1] = "brown"
my_cols[candy$fruity == 1] = "red"
```

```{r, fig.height=15, fig.width=10}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) +
  ylab("")
```

Now, for the first time, using this plot we can answer questions like:

> **Q17. What is the worst ranked chocolate candy?**

```{r}
candy |>
  filter(chocolate ==1) |>
  arrange(winpercent) |>
  head(5)
```

The worst ranked chocolate candy is Sixlets.

> **Q18. What is the best ranked fruity candy?**

```{r}
candy |>
  filter(fruity == 1) |>
  arrange(desc(winpercent)) |>
  head(5)
```

The best ranked fruity candy is Starburst.

## 5.Taking a look at pricepercent

```{r}
library(ggrepel)

# How about a plot of win vs price
ggplot(candy) +
  aes(x = pricepercent, y = winpercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```

> **Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?**

```{r}
candy |>
  arrange(pricepercent, desc(winpercent)) |>
  head(10)
```

The best value candy is Tootsie Pop because it has a relatively high winpercent while having a low pricepercent, meaning it is popular and inexpensive.

> **Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?**

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

The five most expensive candies are Nik L Nip, Nestle Smarties, Ring Pop, Hershey’s Krackel, and Hershey’s Milk Chocolate. Among these, the least popular is Nik L Nip.

#### 5.1 OPTIONAL

```{r, fig.height=15, fig.width=10}
library(ggplot2)

ggplot(candy, aes(x = reorder(rownames(candy), pricepercent),
                  y = pricepercent)) +
  geom_col() +
  coord_flip()
```

Make a lollipop chart of pricepercent

```{r, fig.height=15, fig.width=10}
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col="gray40") +
    geom_point()
```

## 6. Exploring the Correlation Structure

```{r}
library(corrplot)
cij <- cor(candy)
corrplot(cij)
```

> **Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?**

Anti-correlated = negative correlation (red in plot). Chocolate and fruity are anti-correlated, meaning candies that contain chocolate are less likely to be fruity.

> **Q23. Use your corrplot result to make a prediction: which variables do you expect will have the largest contributions (i.e. loadings) to PC1 (i.e., drive the most separation between candies along the first principal component)?**

I expect chocolate and fruity to have the largest contributions to PC1, because they show strong negative correlation and represent two distinct types of candy that separate the dataset.

## 7. Principal Component Analysis

In this case we want to be able to set `scale=TRUE` argument for `prcomp()`because we have one variable, `winpercent`, that is on a very different scale than all others and would otherwise dominate our PCA results.

```{r}
pca <- prcomp(candy, scale = TRUE)
summary(pca)
```

First major result figure is the "score plot" of PC1 vs PC2 - how the different candy are related to each other on our new PC axis:

Now we can plot our main PCA score plot of PC1 vs PC2.

```{r}
plot(pca$x[,1], pca$x[,2])
```

We can change the plotting character and add some color:

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])

p <- ggplot(my_data) +
  aes(x = PC1, 
      y = PC2,
      size = winpercent/100,
      text = rownames(my_data),
      label = rownames(my_data)) +
  geom_point(col = my_cols)

p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

```{r}
ggplot(pca$rotation) +
  aes(x = PC1, y = rownames(pca$rotation)) +
  geom_col()
```

> **Q24. Complete the code to generate the loadings plot above. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you? Where did you see this relationship highlighted previously?**

Fruity, pluribus, and hard load positively on PC1. This makes sense because PC1 separates fruity candies from chocolate candies, consistent with the earlier observed negative correlation between chocolate and fruity.

## 8. Summary

> **Q25. Based on your exploratory analysis, correlation findings, and PCA results, what combination of characteristics appears to make a “winning” candy? How do these different analyses (visualization, correlation, PCA) support or complement each other in reaching this conclusion?**

A “winning” candy is typically chocolate-based, often a bar, and may include caramel or nuts. Visualizations showed chocolate candies rank higher, correlation analysis confirmed chocolate is positively associated with popularity while fruity is negatively associated, and PCA revealed that the main axis of variation separates chocolate from fruity candies. Together, these methods consistently show that chocolate candies are more popular.

### Optional Extension Questions

```{r}
losers = candy[which(candy$winpercent < 50),]
winners = candy[which(candy$winpercent >= 50),]
```

> **Q26. Are popular candies more expensive? In other words: is price significantly different between “winners” and “losers”? List both average values and a P-value along with your answer.**

```{r}
mean(losers$pricepercent)
mean(winners$pricepercent)
```

```{r}
t.test(pricepercent ~ (winpercent >= 50), data = candy)
```

Popular candies are slightly more expensive on average, with winners having a higher mean pricepercent than losers. However, this difference is not statistically significant, as the p-value is greater than 0.05. Therefore, price does not appear to strongly influence candy popularity.

> **Q27. Are candies with more sugar more likely to be popular? What is your interpretation of the means and P-value in this case?**

```{r}
mean(losers$sugarpercent)
mean(winners$sugarpercent)
```

```{r}
t.test(sugarpercent ~ (winpercent >= 50), data = candy)
```

Candies with higher sugar content tend to have slightly higher `winpercent` values on average. However, the difference is not statistically significant, as indicated by a p-value greater than 0.05. This suggests that sugar content alone does not strongly determine candy popularity.
