---
title: "BIMM 143 Lab 6 Hw"
author: 'Mayla (PID: A19136116)'
date: "2026-04-19"
output:
  pdf_document: default
  html_document: default
---

> Section 1.A.

```{r}
normalize <- function(x) {
  (x - min(x, na.rm = TRUE)) / (max(x, na.rm = TRUE) - min(x, na.rm = TRUE))
}
```

```{r}
df <- data.frame(a=1:10, 
                 b=seq(200, 400, length=10), 
                 c=11:20, 
                 d = rep(NA, 10))
                
 df$a <- normalize(df$a)
 df$b <- normalize(df$b)
 df$c <- normalize(df$c)
 df$d <- normalize(df$d)
```
> Section 1.B.

```{r}
# New and Improved code
library(bio3d)
analyze_pdb <- function(pdb_id) {
  pdb <- read.pdb(pdb_id)
  chainA_ca <- trim.pdb(pdb, chain = "A", elety = "CA")
  b_factors <- chainA_ca$atom$b
  plotb3(b_factors, sse = chainA_ca, typ = "l", ylab = "Bfactor")
  return(b_factors)
}
```

```{r}
s1.b <- analyze_pdb("4AKE")
s2.b <- analyze_pdb("1AKE")
s3.b <- analyze_pdb("1E4Y")
```

**Q1. What does your function do?**
This function reads a PDB file, extracts the alpha carbon atoms from chain A, calculates the B-factors, and plots them to visualize protein flexibility. 

**Q2. What does the trim.pdb() function do?**
`trim.pdb()` subsets a PDB structure. In this case, it keeps only chain A and only alpha-carbon (CA) atoms, so the analysis focuses on one comparable backbone trace per protein. 

**Q3. What input parameter would turn off the marginal black and grey rectangles in the plots and what do they represent in this case?**
Set the sse = NULL argument in plotb3() to turn off the marginal black and grey rectangles. In these plots, those rectangles represent secondary structure elements, mainly alpha helices and beta strands/sheets, mapped along the protein sequence. 

**Q4. What would be a better plot to compare across the different proteins?**
A better plot would be a single overlaid line plot with all three B-factor traces on the same axes, or a faceted plot with the same y-axis scale for each protein. That makes direct comparison much easier than looking at three separate plots. 

**Q5. Which proteins are more similar to each other in their B-factor trends. How could you quantify this? HINT: try the rbind(), dist() and hclust() functions together with a resulting dendrogram plot. Look up the documentation to see what each of these functions does:**
```{r}
hc <- hclust( dist( rbind(s1.b, s2.b, s3.b) ) )
plot(hc)
```

The proteins that cluster closest together in the dendrogram have the most similar B-factor trends. To quantify similarity, I can combine the B-factor vectors with `rbind()`, calculate pairwise distances with `dist()`, and cluster them with `hclust()`. Smaller distances indicate more similar B-factor profiles. 

**Q6. How would you generalize the original code above to work with any set of input protein structures?**
I would generalize the code by writing a function that takes any PDB ID as input, reads the structure, trims it to the desired chain/atom type, extracts B-factors, and plots the result. This removes repetition and lets the same analysis work on any protein structure by changing only the function input.
   
### Function Description
   This functions takes a PDB ID as input, reads the protein structure, extracts chain A alpha carbons, and plots B-factor values. 

### Inputs
  - pdb_id: a string representing the PDB ID (e.g. "4AKE")

### Output
  - A plot of B-factor values
  - Returns the B-factor vector
  